WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2014-05-30 ms.tex
Fisher matrix uses conditional expectation value
Fish factors!

\documentclass{iopart}
\usepackage{arydshln}
\usepackage{multirow}
\usepackage{amsopn}
\usepackage{iopams}
\usepackage{graphicx}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\input{ligo-acronyms/acronyms}

@@

The Fisher information matrix for a measurement $x$ described by the unknown parameter vector $\boldsymbol{\theta}$ is
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(x_i ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when---as in this case---the likelihood is Gaussian:
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial h_i}{\partial \theta_j}\right)^*
        \left(\frac{\partial h_i}{\partial \theta_k}\right)
	\right] \frac{1}{S_i(\omega)} \, d\omega.
\end{equation}
%
This form is more useful because it involves only first derivatives, and of the signal $h (\omega)$ rather than the entire observation $x (\omega)$. In terms of the $k$th \ac{SNR}-weighted moment of angular frequency,
%
\begin{equation}\label{eq:angular-frequency-moments}
    {\overline{\omega^k}}_i =
        \left[ \int_0^\infty \frac{|h (\omega)|^2}{S_i(\omega)} \omega^k \, d\omega \right]
        \left[ \int_0^\infty \frac{|h (\omega)|^2}{S_i(\omega)} \, d\omega \right]^{-1},
\end{equation}
%
the Fisher matrix for the signal in the $i$th detector is
%
\begin{equation}
	\mathcal{I}_i = \left(
		\begin{array}{cc}
			\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_i} &
			\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_\mathrm{in}} \\
			{\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_\mathrm{in}}}^\intercal &
			{\rho_i}^2 \mathcal{I}_{\boldsymbol\theta_\mathrm{in},\boldsymbol\theta_\mathrm{in}}
		\end{array}
	\right)
\end{equation}
%
where
%
\begin{equation}\label{eq:fisher-matrix}
	\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_i} = \bordermatrix{
        ~ & \rho_i & \gamma_i & \tau_i \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2 {\overline{\omega}}_i \cr
        \tau_i & 0 & -{\rho_i}^2 {\overline{\omega}}_i & {\rho_i}^2 {\overline{\omega^2}}_i
    }.
\end{equation}
%
To lowest post\nobreakdashes-Newtonian order, the intrinsic parameters affect the phase evolution but not the amplitude evolution, so for BNS systems we can assume $\mathcal{I}_{\rho_i,\boldsymbol\theta_\mathrm{in}} \approx 0$.
%
\marginpar{Show this?}%
%
When we form the Fisher matrix for the whole network, if the detectors all have the same noise \acp{PSD}, $S_1(\omega) = S_1(\omega) = \cdots = S_n(\omega) \equiv S(\omega)$, then we can change variables from phases and times on arrival $\left\{\gamma_i, \tau_i\right\}_i$ to appropriate differences between them, in order to cancel out all off-diagonal terms between the intrinsic and extrinsic parameters. Because all existing ground-based GW detectors have similar bandpasses, we restrict all of our further attention to just the extrinsic part of the Fisher matrix, $\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_i}$.

@@

Fisher matrix elements of \Eref{eq:autocor-likelihood}, with detector subscript suppressed:
%
\begin{eqnarray}
	\nonumber
	\mathcal{I}_{\rho\rho} &=& 1 \\
	\nonumber
	\mathcal{I}_{\rho\gamma} &=& 0 \\
	\nonumber
	\mathcal{I}_{\rho\tau} &=& 0 \\
	\mathcal{I}_{\gamma\gamma} &=& \rho^2
		\int_{-\infty}^\infty \left|a(t)\right|^2 w(t; \rho) dt \\
	\mathcal{I}_{\tau\tau} &=& -\rho^2
		\int_{-\infty}^\infty \Re\left[a^*(t) a''(t)\right] w(t; \rho) dt \\
	\mathcal{I}_{\gamma\tau} &=& -\rho^2
		\int_{-\infty}^\infty \Im\left[a^*(t) a'(t)\right] w(t; \rho) dt \\
	\nonumber\textrm{where} && \\
	\nonumber
	a'(t) &=& \frac{\partial a}{\partial t} \\
	\nonumber
	a''(t) &=& \frac{\partial^2 a}{\partial t^2} \\
	\nonumber
	w(t; \rho) &=& \frac{
		\displaystyle
		\exp\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
		\left(
		I_0\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right] +
		I_1\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
		\right)
	}{
		\displaystyle
		2 \int_{-\infty}^\infty
		\exp\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
		I_0\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
		dt'
	}.
\end{eqnarray}
%
Matrix elements that are not listed have values that are implied by the symmetry of the Fisher matrix. Note that the minus signs are correct but a little confusing: despite them, ${I}_{\gamma\gamma}, {I}_{\tau\tau} \geq 0$ and ${I}_{\gamma\tau} \leq 0$.

Observe that
%
\begin{equation*}
	a(0) = 1,
	\qquad
	a'(0) = i \overline{\omega},
	\qquad
	a''(0) = -\overline{\omega^2}.
\end{equation*}
%
If we can show that the common factor approaches a delta function, $w(t; \rho) \rightarrow \delta(t)$ as $\rho \rightarrow \infty$, then the Fisher matrix approaches \Eref{eq:fisher-matrix} as $\rho \rightarrow \infty$. Certainly it behaves very much like a delta function. The Bessel functions asymptotically approach:
%
\begin{equation*}
	I_0(x), I_1(x) \rightarrow \frac{e^x}{\sqrt{2 \pi x}}
	\textrm{ as } x \rightarrow \infty.
\end{equation*}
%
For large $\rho$, it seems like we could write:
%
\begin{equation*}
	w(t; \rho) \rightarrow \frac{
		\displaystyle
		\exp\left[\frac{\rho^2}{2}|a(t)|^2\right]
	}{
		\displaystyle
		\int_{-\infty}^\infty \exp\left[\frac{\rho^2}{2}|a(t')|^2\right] dt'
	}
	\textrm{ as } \rho \rightarrow \infty.
\end{equation*}

>>>>

\documentclass{iopart}
\usepackage{arydshln}
\usepackage{multirow}
\usepackage{amsopn}
\usepackage{iopams}
\usepackage{graphicx}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\usepackage{protosem}
\input{ligo-acronyms/acronyms}

@@

The Fisher information matrix for a measurement $x$ described by the unknown parameter vector $\boldsymbol{\theta}$ is
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(x_i ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\middle| \boldsymbol\theta
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when---as in this case---the likelihood is Gaussian:
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial h_i}{\partial \theta_j}\right)^*
        \left(\frac{\partial h_i}{\partial \theta_k}\right)
	\right] \frac{1}{S_i(\omega)} \, d\omega.
\end{equation}
%
This form is more useful because it involves only first derivatives, and of the signal $h (\omega)$ rather than the entire observation $x (\omega)$. In terms of the $k$th \ac{SNR}-weighted moment of angular frequency,
%
\begin{equation}\label{eq:angular-frequency-moments}
    {\overline{\omega^k}}_i =
        \left[ \int_0^\infty \frac{|h (\omega)|^2}{S_i(\omega)} \omega^k \, d\omega \right]
        \left[ \int_0^\infty \frac{|h (\omega)|^2}{S_i(\omega)} \, d\omega \right]^{-1},
\end{equation}
%
the Fisher matrix for the signal in the $i$th detector is
%
\begin{equation}
	\mathcal{I}_i = \left(
		\begin{array}{cc}
			\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_i} &
			\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_\mathrm{in}} \\
			{\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_\mathrm{in}}}^\intercal &
			{\rho_i}^2 \mathcal{I}_{\boldsymbol\theta_\mathrm{in},\boldsymbol\theta_\mathrm{in}}
		\end{array}
	\right)
\end{equation}
%
where
%
\begin{equation}\label{eq:fisher-matrix}
	\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_i} = \bordermatrix{
        ~ & \rho_i & \gamma_i & \tau_i \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2 {\overline{\omega}}_i \cr
        \tau_i & 0 & -{\rho_i}^2 {\overline{\omega}}_i & {\rho_i}^2 {\overline{\omega^2}}_i
    }.
\end{equation}
%
To lowest post\nobreakdashes-Newtonian order, the intrinsic parameters affect the phase evolution but not the amplitude evolution, so for BNS systems we can assume $\mathcal{I}_{\rho_i,\boldsymbol\theta_\mathrm{in}} \approx 0$.
%
\marginpar{Show this?}%
%
When we form the Fisher matrix for the whole network, if the detectors all have the same noise \acp{PSD}, $S_1(\omega) = S_1(\omega) = \cdots = S_n(\omega) \equiv S(\omega)$, then we can change variables from phases and times on arrival $\left\{\gamma_i, \tau_i\right\}_i$ to appropriate differences between them, in order to cancel out all off-diagonal terms between the intrinsic and extrinsic parameters. Because all existing ground-based GW detectors have similar bandpasses, we restrict all of our further attention to just the extrinsic part of the Fisher matrix, $\mathcal{I}_{\boldsymbol\theta_i,\boldsymbol\theta_i}$.

@@

We can compute the Fisher matrix elements for the autocorrelation likelihood given by \Eref{eq:autocor-likelihood}, with detector subscript suppressed:
%
\begin{eqnarray}
	\nonumber
	\mathcal{I}_{\rho\rho} &=& 1 \\
	\nonumber
	\mathcal{I}_{\rho\gamma} &=& 0 \\
	\nonumber
	\mathcal{I}_{\rho\tau} &=& 0 \\
	\label{eq:fisher-autocor-gamma-gamma}
	\mathcal{I}_{\gamma\gamma} &=& \rho^2
		\int_{-T}^T \left|a(t)\right|^2 w(t; \rho) dt \\
	\label{eq:fisher-autocor-tau-tau}
	\mathcal{I}_{\tau\tau} &=& -\rho^2
		\int_{-T}^T \Re\left[a^*(t) a''(t)\right] w(t; \rho) dt \\
	\label{eq:fisher-autocor-gamma-tau}
	\mathcal{I}_{\gamma\tau} &=& -\rho^2
		\int_{-T}^T \Im\left[a^*(t) a'(t)\right] w(t; \rho) dt \\
	\nonumber\textrm{where} && \\
	\nonumber
	a'(t) &=& \frac{\partial a}{\partial t} \\
	\nonumber
	a''(t) &=& \frac{\partial^2 a}{\partial t^2} \\
	\nonumber
	w(t; \rho) &=& \frac{
		\displaystyle
		\exp\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
		\left(
		I_0\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right] +
		I_1\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
		\right)
	}{
		\displaystyle
		2 \int_{-T}^T
		\exp\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
		I_0\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
		dt'
	}.
\end{eqnarray}
%
Matrix elements that are not listed have values that are implied by the symmetry of the Fisher matrix. Note that the minus signs are correct but a little confusing: despite them, ${I}_{\gamma\gamma}, {I}_{\tau\tau} \geq 0$ and ${I}_{\gamma\tau} \leq 0$. The time integration limits $[-T, T]$ constitute a flat prior on arrival time, the time coincidence window between detectors.

Observe that
%
\begin{equation*}
	a(0) = 1,
	\qquad
	a'(0) = i \overline{\omega},
	\qquad
	a''(0) = -\overline{\omega^2}.
\end{equation*}
%
If we can show that the common factor approaches a delta function, $w(t; \rho) \rightarrow \delta(t)$ as $\rho \rightarrow \infty$, then the Fisher matrix for the autocorrelation likelihood approaches the Fisher matrix for the full GW data, \Eref{eq:fisher-matrix}, as $\rho \rightarrow \infty$. The Bessel functions asymptotically approach:
%
\begin{equation*}
	I_0(x), I_1(x) \rightarrow \frac{e^x}{\sqrt{2 \pi x}}
	\textrm{ as } x \rightarrow \infty.
\end{equation*}
%
For large $\rho$, the exponential dominates and we can write:
%
\begin{equation*}
	w(t; \rho) \rightarrow \frac{
		\displaystyle
		\exp\left[\frac{\rho^2}{2}|a(t)|^2\right]
	}{
		\displaystyle
		\int_{-T}^T \exp\left[\frac{\rho^2}{2}|a(t')|^2\right] dt'
	}
	\textrm{ as } \rho \rightarrow \infty.
\end{equation*}
%
The Taylor expansion of $|a(t)|^2$ is
%
\begin{equation*}
	|a(t)|^2 = 1 + \frac{1}{2} \left(\frac{\partial^2}{\partial t^2}|a(t)|^2 \Bigg|_{t=0}\right) t^2 + \mathcal{O}(t^4)
	= 1 - {\omega_\mathrm{rms}}^2 t^2 + \mathcal{O}(t^4).
\end{equation*}
%
Substituting, we find that $w(t; \rho)$ approaches a normalized Gaussian distribution:
%
\begin{equation*}
	w(t; \rho) \approx \frac{
		\displaystyle
		\exp\left[-\frac{1}{2} \rho^2 {\omega_\mathrm{rms}}^2 t^2\right]
	}{
		\displaystyle
		\int_{-T}^T \exp\left[-\frac{1}{2} \rho^2 {\omega_\mathrm{rms}}^2 (t')^2\right] dt'
	}.
\end{equation*}
%
And finally, because the Dirac delta function may be defined as the limit of a Gaussian, $w(t; \rho) \rightarrow \delta(t)$ as $\rho \rightarrow \infty$.

We can now write the Fisher matrix for the autocorrelation likelihood in a way that makes a comparison to the full signal model explicit. Define:
%
\begin{eqnarray*}
	\mathcal{I}_\gamma\gamma &=& \rho^2 \cdot \textproto{D}_{\gamma\gamma}(\rho) \\
	\mathcal{I}_\tau\tau &=& \rho^2 \overline{\omega^2} \cdot \textproto{D}_{\tau\tau}(\rho) \\
	\mathcal{I}_\gamma\tau &=& -\rho^2 \overline{\omega} \cdot \textproto{D}_{\gamma\tau}(\rho)
\end{eqnarray*}
%
Now, the $\textproto{D}_{ij}$ %
%
\marginpar{The Fish(er) factor. Go home, Leo.}%
%
contain the integrals from Equations~(\ref{eq:fisher-autocor-gamma-gamma}, \ref{eq:fisher-autocor-tau-tau}, \ref{eq:fisher-autocor-gamma-tau}) and encode the departure of the autocorrelation likelihood from the likelihood of the full data at low SNR. All of the $\textproto{D}_{ij}(\rho)$ are sigmoid-type functions that run from 0 at $\rho = 0$ and asymptotically approach 1 as $\rho \rightarrow \infty$. The transitional SNR $\rho_\mathrm{crit}$ is largely the same for all three nontrivial matrix elements, and is determined by the time coincidence window $T$ and the signal bandwidth $\omega_\mathrm{rms}$.
%
\marginpar{The transitional SNR is probably proportional to some power of $(\omega_\mathrm{rms} T)$.}
